The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-05-03 paper/ms.tex
Small changes throughout

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the thick-disk orbit ({\em top} panel), where the offset in $z$
is much less than $z_{\text{max}}$ of the orbit. The peaks in the distribution
correspond to the turning points of the orbit (or points of maximum vertical
amplitude), where $v_z \sim 0$ and where the star spends most of it orbit.
This is why the distribution is peaked on these values. For the thin-disk
orbit ({\em bottom} panel), the offset in $z$ is comparable to
$z_{\text{max}}$. Now, there will be some points in the orbit where $v_z = 0$
and $z=0$ (in the observed, erroneous coordinate system). At these points, the
computed $J_z$ will vanish. The asymmetry and systematic offset then comes
about because of the constraint that $J_z \geq 0$.\footnote{This is similar to
arguments in cosmology for why gravity produces non-Gaussianity in the density
field, since the density cannot become negative but it can grow arbitrarily
large.}

@@

In the {\em upper middle} panel of Figure~\ref{fig:many_orbit_wrong_ref},
there is essentially no spread in the determination of $J_{\phi}$. This is
expected since, in principle, $J_{\phi}$ only depends on the $x$- and
$y$-components of the position and velocity of the stars\footnote{In practice,
however, $J_{\phi}$ is computed as part of the torus-fitting method.}, and is
thus unaffected by offsets in $z$. Indeed, the result we found in
Figure~\ref{fig:one_orbit_wrong_ref} for the thick-disk orbit holds for all
orbit types.

@@

\begin{deluxetable*}{cccccc}
\tablecaption{Stellar disk scale heights of the real Milky Way and the
simulated galaxies considered in this work.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{cold\tablenotemark{a} gas disk} &
\colhead{stellar thin disk} & \colhead{stellar thick disk} & \colhead{cold gas
softening length} & \colhead{stellar softening length} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
MW\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

@@

\subsection{Description of MW-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-in, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components
are realized with live distributions of N-body particles; the MW and Sgr are
each initialized to be in equilibrium in isolation \Gus{check!}.

The mass resolution of the simulation is $2.6\times10^4$, $1.2\times10^4$, and
$1.0\times10^4$ for the dark matter, disk, and bulge components, respectively.
For the disk and bulge components, a softening length of $30\,\pc$ is used
whereas for the halo a softening length of $60\,\pc$ is used. For Sagittarius,
the softening length for the dark matter and the stars is $60$ and $40\,\pc$,
respectively.

@@

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for the toy simulation of a Sagittarius encounter
\citep{2018MNRAS.481..286L}. Again we have subtracted a best fit curve of the
form $\text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}$, with the values
for $\text{A}$ being $9.5$, $2.5$, $-212$, and $-394\,\pc$, for $\text{B}$
being $0.0013$, $0.0007$, $-0.63$, and $-1.00\,\text{rad}$ and $\text{C}$
being $2.6$, $-7.8$, $-65$, and $-53\,\pc$. Each of these values are for each
successive timestep. In the first two panels, we see that the midplane is
relatively flat in the inner galaxy. But additional encounters drive strong
midplane variation. In the third panel, we see a strong $m=2$ mode develop,
consistent with the $R=8\,\kpc$ panel of Figure~17 in
\citet{2018MNRAS.481..286L}.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels). All calculations are done at
cosmological redshift $\z =0$. The local midplane is determined at a position
$\phi$ by taking the median height of all stars within $R=0.5\,\kpc$ and
$z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma $
error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from previously run toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The {\em upper}
panels show the midplane as a function of azimuth before the first encounter
near the Solar circle. As later encounters occur which bring the
Sagittarius-like object closer to the Solar circle, we can see that stronger
midplane variations result. In fact, one can see that an $m=2$ mode is excited
in the third panel --- this is consistent with Figure~17 of
\citet{2018MNRAS.481..286L}, which shows that an $m=2$ mode is excited (note
that $m=0$ and $m=1$ modes are stronger, but these are removed in our
sine-curve subtraction). The fact that the $t=6.9\,\Gyr$ panel looks
qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that the midplane variations are
produced in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\section{LSR Variations} \label{app:lsr}
We consider the variations of the LSR as a function of azimuth at the fiducial
solar circle ($R_{\odot} = 8.2\,\kpc$). At each azimuth, $\phi$, we take the
median velocity in cylindrical coordinates of all stars within $200\,\pc$ of
the position, following \citet{2018arXiv180610564S}. No best-fit subtraction
was performed as in Figure~\ref{fig:midplane}.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The local standard of rest (LSR) as a function of azimuth at the
fiducial Solar circle ($R_{\odot} = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}).}
\end{figure*}

>>>>

We now describe the intuition behind the shape of Figure~\ref{fig:Jz_hist}.
Consider first the thick-disk orbit ({\em top} panel), where the offset in $z$
is much less than $z_{\text{max}}$ of the orbit. The peaks in the distribution
correspond to the turning points of the orbit (or points of maximum vertical
amplitude), where $v_z \sim 0$ and where the star is on most of it orbit.
This is why the distribution is peaked on these values. For the thin-disk
orbit ({\em bottom} panel), the offset in $z$ is comparable to
$z_{\text{max}}$. Now, there will be some points in the orbit where $v_z = 0$
and $z=0$ (in the observed, erroneous coordinate system). At these points, the
computed $J_z$ will vanish. The asymmetry and systematic offset then comes
about because of the constraint that $J_z \geq 0$.\footnote{This is similar to
arguments in cosmology for why gravity produces non-Gaussianity in the density
field, since the density cannot become negative but it can grow arbitrarily
large.}

@@

In the {\em upper middle} panel of Figure~\ref{fig:many_orbit_wrong_ref},
there is essentially no spread in the determination of $J_{\phi}$. This is
expected since, in principle, $J_{\phi}$ only depends on the $x$- and
$y$-components of the position and velocity of the stars\footnote{In practice,
however, $J_{\phi}$ is computed as part of the torus-fitting method.}, and is
thus unaffected by offsets in $z$. Indeed, the result we found in
Figure~\ref{fig:one_orbit_wrong_ref} for the thick-disk orbit holds for all
orbit types. This is also a demonstration of the robustness of the orbit
integration and action calculation methods we use.

@@

\begin{deluxetable*}{cccccc}
\tablecaption{Stellar and gas disk scale heights of the real Milky Way and the
simulated galaxies considered in this work.\label{tab:scale_height}}
\tablehead{\colhead{galaxy} & \colhead{cold\tablenotemark{a} gas disk} &
\colhead{stellar thin disk} & \colhead{stellar thick disk} & \colhead{cold gas
softening length} & \colhead{stellar softening length} \\
\colhead{} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)} & \colhead{(pc)}
& \colhead{(pc)} }
\startdata
MW\tablenotemark{b} & 40 & 300 & 900 & \nodata & \nodata \\
\mi\tablenotemark{c} & 800\tablenotemark{d} & 480 & 2000 & 12.4 & 11.2 \\
\mf\tablenotemark{c} & 360 & 440 & 1280 & 12.5 & 11.2 \\
\mm\tablenotemark{c} & 250 & 290 & 1030 &10.3 & 11.2 \\
\enddata
% \tablerefs{\citet{2016ARAA..54..529B,2018arXiv180610564S,2008ApJ...673..864J}}
\tablenotetext{a}{$T<1000$ K}
\tablenotetext{b}{\citet{2008ApJ...673..864J,2016ARAA..54..529B}}
\tablenotetext{c}{\citet{2018arXiv180610564S}}
\tablenotetext{d}{The azimuthally averaged gas vertical density profile in
\mi{} is nearly constant to this height, though individual regions show smaller
scale heights and dense clouds.}
\end{deluxetable*}

@@

\subsection{Description of Milky Way-Sagittarius Interaction Simulation} 
\label{ssec:sag_sim}
In addition to the cosmological zoom-ins, we will also briefly consider results
from a toy Sagittarius encounter simulation. This simulation offers us the
ability to see how the midplane varies in a more controlled environment. The
details of the simulation are given in \citet{2018MNRAS.481..286L}, but we
briefly summarize the most relevant details here. For the Milky Way, the dark
halo is modeled as a Hernquist sphere of mass $10^{12}\,\Msun$ and scale
length of $52\,\kpc$, the disk modeled as an exponential disk with a scale
radius of $3.5\,\kpc$, scale height $0.53\,\kpc$, and mass
$6\times10^{10}\,\Msun$, a Hernquist bulge of mass $10^{10}\,\Msun$ and scale
length $0.7\,\kpc$ \citep{1990ApJ...356..359H}. Sagittarius is modelled as a
dark matter Hernquist sphere of mass \Gus{something} and scale length
\Gus{something}, along with a stellar component modeled as a Hernquist sphere
of mass $6.4\times10^8\,\Msun$ and scale radius $0.85\,\kpc$. All components
are realized with live distributions of N-body particles; the MW and Sgr are
each initialized to be in equilibrium in isolation \Gus{check!}.

The mass resolution of the simulation is $2.6\times10^4$, $1.2\times10^4$, and
$1.0\times10^4\,\Msun$ for the dark matter, disk, and bulge components, respectively.
For the disk and bulge components, a softening length of $30\,\pc$ is used
whereas for the halo a softening length of $60\,\pc$ is used. For Sagittarius,
the softening length for the dark matter and the stars is $60$ and $40\,\pc$,
respectively.

The fiducial coordinate system for these toy simulations is the rest frame of
the aligned toy galaxy at the beginning of the simulation.

@@

We compute the same midplane variation in Figure~\ref{fig:midplane_chervin},
but for the toy simulation of a Sagittarius encounter
\citep{2018MNRAS.481..286L}. Again we have subtracted a best fit curve of the
form $\text{A} \sin{\left(\phi + \text{B}\right)} + \text{C}$, with the values
for $\text{A}$ being $9.5$, $2.5$, $-212$, and $-394\,\pc$, for $\text{B}$
being $0.0013$, $0.0007$, $-0.63$, and $-1.00\,\text{rad}$ and $\text{C}$
being $2.6$, $-7.8$, $-65$, and $-53\,\pc$ chronologically for each of the four timesteps shown. In the {\em upper} two panels, we see that the midplane is
relatively flat in the inner galaxy, but additional encounters drive strong
midplane variation. In the {\em lower left} panel, we see a strong $m=2$ mode develop,
consistent with the $R=8\,\kpc$ panel of Figure~17 in
\citet{2018MNRAS.481..286L}.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/midplane_fit.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for the three FIRE galaxies \mi{}, \mf{}, and \mm{} ({\em left},
{\em center}, and {\em right} panels) as a function of azimuthal angle. All calculations are done at
cosmological redshift $\z =0$. The local midplane is determined at a position
$\phi$ by taking the median height of all stars within $R=0.5\,\kpc$ and
$z=1\,\kpc$ (in cylindrical coordinates). The procedure is performed again
using the new height $10$ times to converge on the local midplane height. In
order to allow for the possibility that the fiducial galactocentric coordinate
system is incorrect, we subtract the best fit $A\sin{(\phi+B)}+C$ curve from
each panel --- this figure is reproduced with the original midplane
determination (i.e. before subtracting the best fit sine curve) in
Appendix~\ref{app:lsr}. We then bootstrap resample $1000$ times on all stars
within a $2\,\kpc$ height of the fiducial midplane to determine $1\,\sigma $
error bars, which we report as dashed lines.}
\label{fig:midplane}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from toy simulations
of a Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The {\em upper}
panels show the midplane as a function of azimuth before the first encounter
near the Solar circle. As later encounters occur which bring the
Sagittarius-like object closer to the Solar circle, we can see that stronger
midplane variations result. In fact, one can see that an $m=2$ mode is excited
in the {\em lower left} panel --- this is consistent with Figure~17 of
\citet{2018MNRAS.481..286L}, which shows that an $m=2$ mode is excited (note
that $m=0$ and $m=1$ modes are stronger, but these are removed in our
sine-curve subtraction). The fact that the $t=6.9\,\Gyr$ panel looks
qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) is evidence that the midplane variations are
produced in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\section{LSR Variations} \label{app:lsr}
We consider the variations of the LSR as a function of azimuth at the fiducial
solar circle ($R_{\odot} = 8.2\,\kpc$) in Figure~\ref{fig:lsr_variations}. At
each azimuth, $\phi$, we take the median velocity in cylindrical coordinates
of all stars within $200\,\pc$ of the position, following
\citet{2018arXiv180610564S}. No best-fit subtraction was performed as in
Figure~\ref{fig:midplane}.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/lsr.pdf}
\end{center}
\caption{The local standard of rest (LSR) as a function of azimuth at the
fiducial Solar circle ($R_{\odot} = 8.2\,\kpc$). No best-fit subtraction is
performed here as we did in the case of the midplane
(Section~\ref{ssec:local_midplane}).}
\label{fig:lsr_variations}
\end{figure*}
